Interpolation of Sparse High-Dimensional Data\thanks {This work was supported by the National Science Foundation Grant CNS-1565314.}

### 2020-08-13 [2019-02]_NUMA_(interpolant_error_bound)/tchlux-NA-interpolation/Versions/1-paper.tex
Finished NUMA paper and submitted. Might need to do one more round with editors.

The most challenging problem when scaling in dimension is that the
number of possible interactions between dimensions grows
exponentially. Quantifying all possible interactions becomes
intractable, and hence \changed{problems}{} beyond \changed{ten to
  twenty dimensions usually rely on linear models}{three-dimensional
  data mostly linear models are used}. That is not to say nonlinear
models are absent, but nonlinearities are often either preconceived or
model \changed{interactions between small subsets of}{pairwise
  interactions between} dimensions at most. Even globally nonlinear
approximations such as neural networks are constructed from
compositions of summed low-interaction functions
\cite{clevert2015fast}.

@@

\subsection{Multivariate Adaptive Regression Splines}
\label{sec:mars}
This approximation was introduced in \cite{friedman1991multivariate}
and subsequently improved to its current version in
\cite{stanford1993fast}, called fast multivariate adaptive regression
splines (Fast MARS). In Fast MARS, a least squares fit model is
iteratively built by beginning with a single constant valued function
and adding two new basis functions at each iteration of the form
\begin{align*}
  B_{2j-1}(x) &= B_l(x) \bigl(x_i-x^{(p)}_i\bigr)_+, \\
  B_{2j}(x) &= B_k(x) \bigl(x_i-x^{(p)}_i\bigr)_- ,
\end{align*}
where \changed{$j \leq m$ is the iteration number, $m$ is the maximum
  number of underlying basis functions}{}, $1 \le p \le n$, and
$B_l(x)$, $B_k(x)$ are basis functions from the previous iteration,
 $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$ and
$w_- = (-w)_+$. After iteratively constructing a model, MARS then
iteratively removes basis functions that do not contribute to goodness
of fit. In effect, MARS creates a locally component-wise tensor
product approximation of the data. The overall computational
complexity of Fast MARS is $\mathcal{O}(n d m^3)$ \changed{}{where $m$
  is the maximum number of underlying basis functions}. This paper
uses an implementation of Fast MARS \cite{rudy2017pyearth} with $m =
200$ throughout all experiments.

@@

\subsection{Multilayer Perceptron Regressor}
\label{sec:mlp}
The neural network is a well studied and widely used method for both
regression and classification tasks
\cite{hornik1989multilayer,rumelhart1988learning}. When using the
rectified linear unit (ReLU) activation function
\cite{dahl2013improving} and fitting the model with stochastic
gradient descent (SGD) or BFGS minimization techniques
\cite{goh2017why,moller1993scaled,robbins1951stochastic}, the model
built by a multilayer perceptron uses layers $l : \mathbb{R}^{i}
\rightarrow \mathbb{R}^{j}$ defined by
 $$ l(u) = \bigl( u^t W_l + c_l \bigr)_+ ,$$
where $c_l \in \mathbb{R}^j$ and $W_l$ is the $i \times j$ weight
matrix for layer $l$. In this form, the multilayer perceptron (MLP)
produces a piecewise linear model of the input data. The computational
complexity of training a multilayer perceptron is $\mathcal{O}(n d
m)$, where $m$ is determined by the sizes of the layers of the network
and the stopping criterion of the minimization used for finding
weights. This paper uses an MLP built from Keras and Tensorflow to
perform regression \cite{tensorflow2015-whitepaper,chollet2015keras}
with ten hidden layers each having thirty two nodes (a total of
approximately ten thousand parameters), ReLU activation, and
\changed{one hundred}{five} thousand \changed{steps}{epochs} of SGD
for training. \changed{It should be noted that the choice of neural
  network architecture impacts approximation performance, but no
  architecture tuning is performed here}{}.

@@

The modified Shepard method used here (ShepMod) generates a continuous
approximation based on Euclidean distance and resembles a nearest
neighbor interpolant \cite{cover1967nearest}. This model is a type of
\textit{metric interpolation}, also called a Shepard method
\cite{gordon1978shepard,shepard1968two}. The interpolant has the form
 $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
     {\sum\limits_{k=1}^{n}W_k(x)} ,$$
where \changed{$W_k(x^{(k)}) = 1$, $W_k(x) = \bigl(\bigl(r_k -
  \bigl\|x - x^{(k)}\bigr\|_2\bigr)_+ \big/ \bigl( r_k \bigl\|x -
  x^{(k)}\bigr\|_2 \bigr)\bigr)^2$ for $x \neq x^{(k)}$,}{} and $r_k
\in \mathbb{R}$ is the smallest radius such that at least $d+1$ other
points $x^{(j)}$, $j \not = k$, are inside the closed Euclidean ball
of radius $r_k$ about $x^{(k)}$. The computational complexity of
ShepMod is $\mathcal{O}(n^2d)$. This paper uses a Fortran 95
implementation of ShepMod derived from SHEPPACK
\cite{thacker2010algorithm}.

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth,]{example_shepard_with_weight.pdf}
  \caption{\color{red} On the left above is a depiction of the radius
    of influence for three chosen points of a collection in two
    dimensions using the modified Shepard criteria. On the right is
    the shape of the weight function $W_k$ for all data points, where
    $W_k(x) = 0$ when $\|x-x^{(k)}\|_2 \geq r_k$. It can also be seen
    that in the limit as $W_k$ tends towards infinity at $x_k$, the
    relative weight of $x_k$ tends to one.
  \vspace{-.1cm}}
  \label{fig:shepard}
\end{figure}
}{}

\subsection{Linear Shepard}
The linear Shepard method (LSHEP) is a blending function using local
linear interpolants, a special case of the general Shepard algorithm
\cite{thacker2010algorithm}. The interpolant has the form
 $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}
 {\sum\limits_{k=1}^{n}W_k(x)} ,$$
where $W_k(x)$ is \changed{the same as for the modified Shepard
  method}{a locally supported weighting function} and $P_k(x)$ is a
local linear approximation to the data satisfying
$P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$. The computational
complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the
Fortran 95 implementation of LSHEP in SHEPPACK
\cite{thacker2010algorithm}.

@@

The box spline model used here is \changed{an}{sparse grid}
interpolation technique built from overlapping box splines
\cite{de2013box}. The box splines serve as basis functions that are
shifted and scaled to have support over box shaped regions. The boxes
are constructed in a way to guarantee a covering of the domain
\cite{lux2018novel}. Given a set of box splines
$\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties outlined
in \cite{lux2018novel} and anchored at interior points
$\bigl\{x^{(i)}\bigr\}$,
 $$ \hat f(y) = \frac{\sum\limits_{i=1}^n b^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n b^{x^{(i)}}(y)}. $$
Note that the box splines always satisfy
$b^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $b^{x^{(i)}}(y)
\geq 0$. The computational complexity of interpolation via the box
spline model is $\mathcal{O}(n^2 d)$.

@@

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth,]{example_distribution_prediction.pdf}
  \caption{In this HPC I/O example, the general methodology for
    predicting a CDF and evaluating error can be seen\color{red},
    where M means $\times 10^6$\color{black}. The Delaunay method
    chose three source distributions (dotted lines) and assigned
    weights \color{red}\{.1, .3, .6\} \color{black} (top to bottom at
    \color{red}middle\color{black}). The weighted sum of the three
    known CDFs produces the predicted CDF (dashed line). The KS
    Statistic (\color{red}vertical line\color{black}) computed between
    the true CDF (solid line) and predicted CDF (dashed line) is 0.2
    for this example. For this example the KS test null hypothesis is
    rejected at $p$-value 0.01, however it is not rejected at
    $p$-value 0.001.
  \vspace{-.1cm}}
  \label{fig:prediction-example}
\end{figure}
}{}

@@

The two-sample KS test is a useful nonparametric test for comparing
two \changed{empirical}{} CDFs while only assuming stationarity,
finite mean, and finite variance. The null hypothesis (that two
\changed{empirical}{} CDFs come from the same underlying distribution)
is rejected at level $p \in [0,1]$ when
 $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)}
\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$ with distribution sample
sizes $n_1,n_2 \in \mathcal{N}$. For all applications of the KS test
presented in this work $n_1 = n_2$. An example of \changed{the process
  of generating predicted distributions from known distributions and
  the subsequent}{the round-trip prediction methodology from known and
  predicted distributions to the} calculation of error can be seen in
Figure \ref{fig:prediction-example}. A brief listing of relevant
statistical terms used throughout this work is provided in the
Appendix (Section \ref{sec:appendix}).

@@

%% \newpage

@@

\begin{proofdot}
  \def\g{\gamma_g}
  Consider $g(t)=f(x(t)) - \hat f(x(t))$ along
  the line segment $x(t)=(1-t)x_0 + tx_j$, $0\le t\le1$, from $x_0$ to
  $x_j$.  Observe that $g(0)=f(x_0) - \hat f(x_0) = 0$, $g(1)=f(x_j) -
  \hat f(x_j) = 0$, and $g'(t)$ is $\g$-Lipschitz continuous with $\g
  = \gamma \|x_j-x_0\|_2^2$. \changed{The following is visualized in 
  Figure \ref{fig:lemma3}.}{}

  Suppose $g'(0) > \g/2$. Then $|g'(0)-g'(t)| \le \g t \implies g'(0)
  - \g t \le g'(t)$, and the line $w=g'(0)-\g t$ intersects the
  $t$-axis at $\tilde t=g'(0)/\g > 1/2$. $\tilde t<1$ necessarily,
  since by Rolle's Theorem there exists $0<z<1$ such that $g'(z)=0$
  and so $g'(0) - \g z \le g'(z) = 0$. Now \changed{by integrating}{}
  $$\frac{g'(0)^2}{2\g} = \frac{g'(0)\tilde t}{2} = \int_0^{\tilde t}
  (g'(0)-\g t)dt \le \int_0^{\tilde t} g'(t)dt = g(\tilde t),$$ and
  \changed{acknowledging $g'(0) > \g / 2$}{}
  \begin{align*}
    \frac{-g'(0)^2}{2\g} < g'(0) - \frac{g'(0)^2}{2\g} - \g/2 =
    \frac{(1-\tilde t)(g'(0)-\g)}{2} &= \int_{\tilde t}^1 (g'(0)-\g
    t)dt \\ &\le \int_{\tilde t}^1 g'(t)dt = -g(\tilde t),
  \end{align*}
  \noindent a contradiction for the value of $g(\tilde t)$. A similar
  contradiction arises for $g'(0)<-\g/2$. Therefore $|g'(0)| \le
  \g/2$. In terms of $f$,
  $$\bigl|\langle \nabla f(x_0)-c,x_j-x_0\rangle\bigr| = |g'(0)| \le
  \gamma \|x_j-x_0\|_2^2/2 \le \gamma k^2 / 2,$$ which holds for all
  $1\le j\le d$.  Finally, using {\it Lemma \ref{lemma:2}},
  $$\|\nabla f(x_0)-c\|_2^2 \le \frac{d}{\sigma_d^2}(\gamma k^2/2)^2
  \implies \|\nabla f(x_0)-c\|_2 \le \frac{\sqrt{d} \, \gamma
    k^2}{2\sigma_d}.$$ \qed
\end{proofdot}

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth,]{example_lemma_3.pdf}
  \def\g{\gamma_g}
  \caption{\color{red} Three different scenarios visualizing {\it
      Lemma \ref{lemma:3}}, where $g(t)$ is the difference between a
    piecewise linear interpolant and the approximated function along a
    normalized line segment between interpolation points, $g'(t)$ is
    $\g$-Lipschitz continuous, and $w$ and $\tilde t$ are defined in
    the proof. Leftmost is a randomly chosen permissible shape of $g$
    and $g'$. The middle is the only possible shape of $g$ and $g'$
    given $g'(0) = \g / 2$, establishing the case of equality in the
    lemma. Rightmost is the resulting contradiction when $g'(0) > \g /
    2$, notice it is impossible to ensure $g'(t)$ is $\g$-Lipschitz
    continuous and satisfy $g(1) = 0$ (highlighted with red circle on
    the right).
  \vspace{-.1cm}}
  \label{fig:lemma3}
\end{figure}
}{}

@@

In summary, the approximation error of a linear (simplicial)
interpolant tends quadratically towards zero when approaching observed
data only when the diameter of the simplex is also reduced at a
proportional rate. Only linear convergence to the true function
\changed{is}{can be} achieved in practice\changed{}{,} without the
incorporation of additional observations\changed{, by moving closer to
  interpolation points}{}. Notice that the approximation error is
largely determined by the spacing of observed data. Predictions made
by simplices whose vertices are not well-spaced (i.e., have large
diameter, or are nearly contained in a hyperplane) have higher error.

@@

The theoretical results constructed in Section \ref{sec:theory} for
(piecewise) linear interpolation are promising and apply directly to
Delaunay interpolation, however they are difficult to interpret in
context with approximation algorithms that do not have similar known
uniform error bounds. For that reason, an analytic function is used to
measure the error of a piecewise linear interpolant (Delaunay) and a
piecewise linear regressor (the MLP) when provided an increasing
amount of data with varying levels of random noise. \changed{Only
  Delaunay and the MLP are considered in this demonstration because
  all other mentioned techniques are not strictly piecewise linear
  approximations.}{}

@@

The error of a linear interpolant constructed from well spaced points
is dominated by the distance to the nearest data point. In order to
uniformly decrease the distance to the nearest data point across the
unit cube, exponentially more data points must be available for
approximation. For this experiment $N = 2(4^i),$ $i = 1,$ $\ldots,$
$6,$ chosen points are kept well spaced by computing approximate
Fekete points. \changed{The following experiment was also run with a
  Latin hypercube design \cite{park1994optimal}, which produced almost
  identical results that are not reported here.}{} Fekete points have
a history in potential theory \cite{kovari_pommerenke_1968} and are
most generally defined as those points that maximize the absolute
value of a Vandermonde determinant. Here, the QR method outlined in
\cite{bos2010computing} is used to identify approximate Fekete points
from a Vandermonde matrix. A multivariate polynomial basis of size
$2N$ is constructed containing all polynomials of degree $\leq n$ for
the largest $n$ such that ${d+n \choose n} \leq 2N$ and $2N - {d+n
  \choose n}$ arbitrarily selected polynomials of degree $n + 1.$ The
Vandermonde matrix for these $2N$ polynomials is used to select $N$
approximate Fekete points.

Finally an additional aspect is added to this test problem by
incorporating random uniform noise into the evaluations of $f.$ For
each test two experiments are executed, one with exact function
evaluations (\changed{an interpolation problem}{}) and one with a
constant signal-to-noise ratio (SNR) of 10:1 (\changed{a regression
  problem}{}).

@@

The bound from the theorem suggests that \changed{by}{exponentially}
increasing the number of \changed{well-spaced}{} data points,
\changed{approximation error can be reliably decreased}{results in a
  decreasing error}. The $d = 2$ test seen on the left half of Figure
\ref{fig:convergence-2d} shows a consistent decrease in error for
Delaunay and also shows the eventual accuracy plateau obtained by a
parametric regression form (the MLP, at roughly 500 points). On the
right hand side of Figure \ref{fig:convergence-2d} the random noise
clearly prohibits Delaunay from converging to $f$, while the MLP is
able to improve its approximation with more data points on average.
The convergence result for a very low-dimensional problem like this is
expected.  However, intuition fails for higher dimensional problems.

@@

\changed{
\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.9\textwidth,]{theorem-convergence.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Number of Data Points};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Error, Distance, and Smallest SV };
  \end{tikzpicture}
  \caption{\color{red}The distribution of absolute error, distance to
    the nearest data point, smallest singular value (SV) and the
    longest edge of the simplex containing each approximation point in
    the tests from Fig. \ref{fig:convergence-2d} and
    Fig. \ref{fig:convergence-20d} for Delaunay. In two dimensions it
    can be seen that $\|z - x_0\|_2$, $\sigma_d$, and $k$ all shrink
    at the same rate for well-spaced approximation points, resulting
    in a faster rate of decrease for approximation error. Notice that
    in higher dimension the data remains sparse even with thousands of
    data points, and the decay in data spacing is more prominant. The
    relatively small reduction in $k$ along with the decrease in
    $\sigma_d$ explain the minimal reduction in error seen by Delaunay
    in Fig. \ref{fig:convergence-20d}.}
  \label{fig:data-spacing}
\end{figure}
}{}

\changed{The results achieved for both $d = 2$ and $d = 20$ align with
  the theoretical error bound as can be seen in Figure
  \ref{fig:data-spacing}. In low dimension, thousands of data points
  meaningfully reduce the sparsity of the approximation problem.
  However, in higher dimension it takes many more points to achieve a
  reduction in data sparsity while simultaneously being more difficult
  to produce well-\color{red}conditioned simplices. This evidences the inherent
  challenge of data sparsity in high dimension approximation
  problems. The analytic results presented are not caused by the
  chosen test function, but rather the exponential increase in
  complexity that accompanies increased dimension.}{}

In summary, the regime of \changed{signal-to-noise ratios that 
  result in competitively accurate interpolants}{} is \changed{}{much}
greater for moderate to high dimensional problems \changed{due to
  increased sparsity}{, even for functions with random noise
  incorporated into their evaluation}. Acknowledging the viability of
interpolation for problems of moderate dimension, the next section
will consider real-world problems of similar proportion (thousands of
examples in tens of dimensions).

@@

In addition to the figures displaying approximation results for each
data set, tables of accompanying numerical results \changed{including
  timings and exact prediction errors}{} are located in the Appendix
(Section \ref{sec:appendix}). All of the test data sets capture
underlying functions that are almost certainly stochastic. As
described in Section \ref{sec:introduction}, regression techniques
appear most appropriate for these problems. However, typically data
grows exponentially more sparse with increasing dimension. Given that
sparse data regressors tend towards interpolation and, as demonstrated
in Section \ref{sec:analytic}, interpolants produce similar (if not
identical) results, it is presumed that interpolants are equally
viable approximation techniques on these problems.

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig4.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Histogram of forest fire area burned under recorded weather
    conditions. The data is presented on a $\ln$ scale because
    most values are small with exponentially fewer fires on record
    that burn large areas.}
  \label{fig:hist-forest-fire}

  \vspace{.3cm}

  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig5.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Area Burned Error};
  \end{tikzpicture}
  \caption{All models are applied to approximate the amount of area
    that would be burned given environment conditions. $10$-fold cross
    validation as described in the beginning of Section \ref{sec:data}
    is used to evaluated each algorithm. This results in exactly one
    prediction from each algorithm for each data point. These boxes
    depict the median (middle bar), median $95\%$ confidence interval
    (cones), quartiles (box edges), fences at $3/2$ interquartile
    range (whiskers), and outliers (dots) of absolute prediction error
    for each model. Similar to Figure \ref{fig:hist-forest-fire}, the
    errors are presented on a $\ln$ scale. The numerical data
    corresponding to this figure is provided in Table
    \ref{table:error-forest-fire} in the Appendix.}
  \label{fig:error-forest-fire}
\end{figure}

>>>>

The most challenging problem when scaling in dimension is that the
number of possible interactions between dimensions grows
exponentially. Quantifying all possible interactions becomes
intractable, and hence beyond three-dimensional data mostly linear
models are used. That is not to say nonlinear models are absent, but
nonlinearities are often either preconceived or model pairwise
interactions between dimensions at most. Even globally nonlinear
approximations such as neural networks are constructed from
compositions of summed low-interaction functions
\cite{clevert2015fast}.

@@

\subsection{Multivariate Adaptive Regression Splines}
\label{sec:mars}
This approximation was introduced in \cite{friedman1991multivariate}
and subsequently improved to its current version in
\cite{stanford1993fast}, called fast multivariate adaptive regression
splines (Fast MARS). In Fast MARS, a least squares fit model is
iteratively built by beginning with a single constant valued function
and adding two new basis functions at each iteration of the form
\begin{align*}
  B_{2j-1}(x) &= B_l(x) \bigl(x_i-x^{(p)}_i\bigr)_+, \\
  B_{2j}(x) &= B_k(x) \bigl(x_i-x^{(p)}_i\bigr)_- ,
\end{align*}
where $j$ is the iteration number, $1 \le p \le n$, and $B_l(x)$,
$B_k(x)$ are basis functions from the previous iteration,
 $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$
and $w_- = (-w)_+$. After iteratively constructing a model, MARS then
iteratively removes basis functions that do not contribute to goodness
of fit. In effect, MARS creates a locally component-wise tensor
product approximation of the data. The overall computational
complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is the
maximum number of underlying basis functions. This paper uses an
implementation of Fast MARS \cite{rudy2017pyearth} with $m = 200$
throughout all experiments.

@@

\subsection{Multilayer Perceptron Regressor}
\label{sec:mlp}
The neural network is a well studied and widely used method for both
regression and classification tasks
\cite{hornik1989multilayer,rumelhart1988learning}. When using the
rectified linear unit (ReLU) activation function
\cite{dahl2013improving} and fitting the model with stochastic
gradient descent (SGD) or BFGS minimization techniques
\cite{goh2017why,moller1993scaled,robbins1951stochastic}, the model
built by a multilayer perceptron uses layers $l : \mathbb{R}^{i}
\rightarrow \mathbb{R}^{j}$ defined by
 $$ l(u) = \bigl( u^t W_l + c_l \bigr)_+ ,$$
where $c_l \in \mathbb{R}^j$ and $W_l$ is the $i \times j$ weight
matrix for layer $l$. In this form, the multilayer perceptron (MLP)
produces a piecewise linear model of the input data. The computational
complexity of training a multilayer perceptron is $\mathcal{O}(n d
m)$, where $m$ is determined by the sizes of the layers of the network
and the stopping criterion of the minimization used for finding
weights. This paper uses an MLP built from Keras and Tensorflow to
perform regression \cite{tensorflow2015-whitepaper,chollet2015keras}
with ten hidden layers each having thirty two nodes (a total of
approximately ten thousand parameters), ReLU activation, and five
thousand epochs of SGD for training.

@@

The modified Shepard method used here (ShepMod) generates a continuous
approximation based on Euclidean distance and resembles a nearest
neighbor interpolant \cite{cover1967nearest}. This model is a type of
\textit{metric interpolation}, also called a Shepard method
\cite{gordon1978shepard,shepard1968two}. The interpolant has the form
 $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
     {\sum\limits_{k=1}^{n}W_k(x)} ,$$
where $W_k(x) = \bigl(r_k - \bigl\|x - x^{(k)}\bigr\|\bigr)_+^2 \big/
\bigl\|x - x^{(k)}\bigr\|_2^2$ and $r_k \in \mathbb{R}$ is the
smallest radius such that at least $d+1$ other points $x^{(j)}$, $j
\not = k$, are inside the closed Euclidean ball of radius $r_k$ about
$x^{(k)}$. The computational complexity of ShepMod is
$\mathcal{O}(n^2d)$. This paper uses a Fortran 95 implementation of
ShepMod derived from SHEPPACK \cite{thacker2010algorithm}.

\subsection{Linear Shepard}
The linear Shepard method (LSHEP) is a blending function using local
linear interpolants, a special case of the general Shepard algorithm
\cite{thacker2010algorithm}. The interpolant has the form
 $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}
 {\sum\limits_{k=1}^{n}W_k(x)} ,$$
where $W_k(x)$ is a locally supported weighting function and $P_k(x)$
is a local linear approximation to the data satisfying
$P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$. The computational
complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the
Fortran 95 implementation of LSHEP in SHEPPACK
\cite{thacker2010algorithm}.

@@

The box spline model used here is a sparse grid interpolation
technique built from overlapping box splines \cite{de2013box}. The box
splines serve as basis functions that are shifted and scaled to have
support over box shaped regions. The boxes are constructed in a way to
guarantee a covering of the domain \cite{lux2018novel}. Given a set of
box splines $\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box
properties outlined in \cite{lux2018novel} and anchored at interior
points $\bigl\{x^{(i)}\bigr\}$,
 $$ \hat f(y) = \frac{\sum\limits_{i=1}^n b^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n b^{x^{(i)}}(y)}. $$
Note that the box splines always satisfy
$b^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $b^{x^{(i)}}(y)
\geq 0$. The computational complexity of interpolation via the box
spline model is $\mathcal{O}(n^2 d)$.

@@

\begin{figure}[htb]
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig1.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
  \end{tikzpicture}
  \vspace{-0.3cm}
  \caption{In this HPC I/O example, the general methodology for
    predicting a CDF and evaluating error can be seen. The Delaunay
    method chose three source distributions (dotted lines) and
    assigned weights \{.3, .4, .3\} (top to bottom at arrow). The
    weighted sum of the three known CDFs produces the predicted CDF
    (dashed line). The KS Statistic (arrow) computed between the true
    CDF (solid line) and predicted CDF (dashed line) is 0.2 for this
    example. For this example the KS test null hypothesis is rejected
    at $p$-value 0.01, however it is not rejected at $p$-value 0.001.
  \vspace{-.1cm}}
  \label{fig:prediction-example}
\end{figure}

@@

The two-sample KS test is a useful nonparametric test for comparing
two CDFs while only assuming stationarity, finite mean, and finite
variance. The null hypothesis (that two CDFs come from the same
underlying distribution) is rejected at level $p \in [0,1]$ when
 $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all
applications of the KS test presented in this work $n_1 = n_2$. An
example of the round-trip prediction methodology from known and
predicted distributions to the calculation of error can be seen in
Figure \ref{fig:prediction-example}. A brief listing of relevant
statistical terms used throughout this work is provided in
the Appendix (Section \ref{sec:appendix}).

@@

@@

\begin{proofdot}
  \def\g{\gamma_g}
  Consider $g(t)=f(x(t)) - \hat f(x(t))$ along
  the line segment $x(t)=(1-t)x_0 + tx_j$, $0\le t\le1$, from $x_0$ to
  $x_j$.  Observe that $g(0)=f(x_0) - \hat f(x_0) = 0$, $g(1)=f(x_j) -
  \hat f(x_j) = 0$, and $g'(t)$ is $\g$-Lipschitz continuous with $\g
  = \gamma \|x_j-x_0\|_2^2$.

  Suppose $g'(0) > \g/2$. Then $|g'(0)-g'(t)| \le \g t \implies g'(0)
  - \g t \le g'(t)$, and the line $w=g'(0)-\g t$ intersects the
  $t$-axis at $\tilde t=g'(0)/\g > 1/2$. $\tilde t<1$ necessarily,
  since by Rolle's Theorem there exists $0<z<1$ such that $g'(z)=0$
  and so $g'(0) - \g z \le g'(z) = 0$. Now
  $$\frac{g'(0)^2}{2\g} = \frac{g'(0)\tilde t}{2} = \int_0^{\tilde t}
  (g'(0)-\g t)dt \le \int_0^{\tilde t} g'(t)dt = g(\tilde t),$$ and
  \begin{align*}
    \frac{-g'(0)^2}{2\g} < g'(0) - \frac{g'(0)^2}{2\g} - \g/2 =
    \frac{(1-\tilde t)(g'(0)-\g)}{2} &= \int_{\tilde t}^1 (g'(0)-\g
    t)dt \\ &\le \int_{\tilde t}^1 g'(t)dt = -g(\tilde t),
  \end{align*}
  \noindent a contradiction for the value of $g(\tilde t)$. A similar
  contradiction arises for $g'(0)<-\g/2$. Therefore $|g'(0)| \le
  \g/2$. In terms of $f$,
  $$\bigl|\langle \nabla f(x_0)-c,x_j-x_0\rangle\bigr| = |g'(0)| \le
  \gamma \|x_j-x_0\|_2^2/2 \le \gamma k^2 / 2,$$ which holds for all
  $1\le j\le d$.  Finally, using {\it Lemma \ref{lemma:2}},
  $$\|\nabla f(x_0)-c\|_2^2 \le \frac{d}{\sigma_d^2}(\gamma k^2/2)^2
  \implies \|\nabla f(x_0)-c\|_2 \le \frac{\sqrt{d} \, \gamma
    k^2}{2\sigma_d}.$$ \qed
\end{proofdot}

@@

In summary, the approximation error of a linear (simplicial)
interpolant tends quadratically towards zero when approaching observed
data only when the diameter of the simplex is also reduced at a
proportional rate.  Only linear convergence to the true function can
be achieved in practice, without the incorporation of additional
observations. Notice that the approximation error is largely
determined by the spacing of observed data. Predictions made by
simplices whose vertices are not well-spaced (i.e., have large
diameter, or are nearly contained in a hyperplane) have higher
error.

@@

The theoretical results constructed in Section \ref{sec:theory} for
(piecewise) linear interpolation are promising and apply directly to
Delaunay interpolation, however they are difficult to interpret in
context with approximation algorithms that do not have similar known
uniform error bounds. For that reason, an analytic function is used to
measure the error of a piecewise linear interpolant (Delaunay) and a
piecewise linear regressor (the MLP) when provided an increasing
amount of data with varying levels of random noise.

@@

The error of a linear interpolant constructed from well spaced points
is dominated by the distance to the nearest data point. In order to
uniformly decrease the distance to the nearest data point across the
unit cube, exponentially more data points must be available for
approximation. For this experiment $N = 2(4^i),$ $i = 1,$ $\ldots,$
$6,$ chosen points are kept well spaced by computing approximate
Fekete points. Fekete points have a history in potential theory
\cite{kovari_pommerenke_1968} and are most generally defined as those
points that maximize the absolute value of a Vandermonde
determinant. Here, the QR method outlined in \cite{bos2010computing}
is used to identify approximate Fekete points from a Vandermonde
matrix. A multivariate polynomial basis of size $2N$ is constructed
containing all polynomials of degree $\leq n$ for the largest $n$ such
that ${d+n \choose n} \leq 2N$ and $2N - {d+n \choose n}$
arbitrarily selected polynomials of degree $n + 1.$ The Vandermonde
matrix for these $2N$ polynomials is used to select $N$ approximate
Fekete points.

Finally an additional aspect is added to this test problem by
incorporating random uniform noise into the evaluations of $f.$ For
each test two experiments are executed, one with exact function
evaluations (for interpolation) and one with a constant signal-to-noise
(SNR) ratio of 10:1 (for regression).

@@

The bound from the theorem suggests that exponentially increasing the
number of data points should result in a decreasing error. The $d = 2$
test seen on the left half of Figure \ref{fig:convergence-2d} shows a
consistent decrease in error for Delaunay and also shows the eventual
accuracy plateau obtained by a parametric regression form (the MLP, at
roughly 500 points). On the right hand side of Figure
\ref{fig:convergence-2d} the random noise clearly prohibits Delaunay
from converging to $f$, while the MLP is able to improve its
approximation with more data points on average.  The convergence
result for a very low-dimensional problem like this is expected.
However, intuition fails for higher dimensional problems.

@@

In summary, the regime of interpolation is much greater for moderate
to high dimensional problems, even for functions with random noise
incorporated into their evaluation. Acknowledging the viability of
interpolation for problems of moderate dimension, the next section
will consider real-world problems of similar proportion (thousands of
examples in tens of dimensions).

@@

In addition to the figures displaying approximation results for each
data set, tables of accompanying numerical results are located in the
Appendix (Section \ref{sec:appendix}). All of the test data sets
capture underlying functions that are almost certainly stochastic. As
described in Section \ref{sec:introduction}, regression techniques
appear most appropriate for these problems. However, typically data
grows exponentially more sparse with increasing dimension. Given that
sparse data regressors tend towards interpolation and, as demonstrated
in Section \ref{sec:analytic}, interpolants produce similar (if not
identical) results, it is presumed that interpolants are equally
viable approximation techniques on these problems.

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig4.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Forest Fire Area Burned};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Histogram of forest fire area burned under recorded weather
    conditions. The data is presented on a $\ln$ scale because
    most values are small with exponentially fewer fires on record
    that burn large areas.}
  \label{fig:hist-forest-fire}
\end{figure}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.8\textwidth,]{Fig5.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Area Burned Error};
  \end{tikzpicture}
  \caption{All models are applied to approximate the amount of area
    that would be burned given environment conditions. $10$-fold cross
    validation as described in the beginning of Section \ref{sec:data}
    is used to evaluated each algorithm. This results in exactly one
    prediction from each algorithm for each data point. These boxes
    depict the median (middle bar), median $95\%$ confidence interval
    (cones), quartiles (box edges), fences at $3/2$ interquartile
    range (whiskers), and outliers (dots) of absolute prediction error
    for each model. Similar to Figure \ref{fig:hist-forest-fire}, the
    errors are presented on a $\ln$ scale. The numerical data
    corresponding to this figure is provided in Table
    \ref{table:error-forest-fire} in the Appendix.}
  \label{fig:error-forest-fire}
\end{figure}
